--- 
title: "TMA4315: Compulsory exercise 1 (title)" 
subtitle: "Group 0: Name1, Name2 (subtitle)" 
date: "`r format(Sys.time(), '%d.%m.%Y')`" # the current date, can be regular text as well
output: # 3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  # pdf_document:
  #  toc: false
  #  toc_depth: 2

---

```{r setup, include = FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)
```

# Part 1

**Bold**

_italic_

To get a pdf file, make comments of the lines with the "html_document" information, and make the lines with the "pdf_document" information regular, and vice versa.

## a)

Your answer for part 1a)

```{r setup, include=FALSE}
install.packages("car")
library(car)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID), ]

```

```{r}
# some R code for part 1a)
library(GGally)
ggpairs(SLID, lower = list(combo = wrap(ggally_facethist, binwidth = 0.5)))
```

From the top row we can see that there is a noticeable corrolation between wages, and the education, age and sex. Language on the other hand doesn't seem to have a large impact on the wages. One can see that people with high level (20) of education are distributed over the whole span of wages, but low educated people are centred around low wages, with very few or none at high level of wages.

The age corrolates to wages in that there is people from all age categories that have low wages, but the there is more likley to have a higher wage around age 40, and decreasing when younger or older.


There is a corrolation between age and education, in that the education level decreases as the age increases. There is a known fact that the average education level have increased over the last 50 years, which corrolates with the data set. 


First we assume that there is a linear relationship between the covariates, ie. the relationship can be expressed as:
$$\bf{Y} = \bf{X} \bf{\beta} + \bf{\varepsilon}$$

Linearity of covariates: Y=Xβ+ε. Problem: non-linear relationship?

Homoscedastic error variance: Cov(ε)=σ2I. Problem: Non-constant variance of error terms

Uncorrelated errors: Cov(εi,εj)=0.

Additivity of errors: Y=Xβ+ε

Assumption of normality: ε∼Nn(0,σ2I)

# Part 2

**Bold**

_italic_



```{r}
# some R code for part 2a)
library(mylm)
model1 <- mylm(wages ~ education, data = SLID)
print(model1)
model1b <- lm(wages ~ education, data = SLID)
print(model1b)

```
```{r}
# some R code for part 2b)
summary(model1)
summary(model1b)

```




